---
title: "Peer Assessment 2 - Analysis of Storm Data"
author: "Kevin Gardner"
date: "02/14/2016"
output: html_document
---

## Synopsis  
This document contains R code and documentation to prepare and analyze the following dataset `https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2`.  

Our analysis answers two main questions:  

* Which types of events are most harmful to population health?  
* Which types of events have the greatest economic consequences?  

The report contains three plots which illustrate the results of the analysis.  

## Data processing  
Setting up environment  
```{r, tidy=TRUE, hide=TRUE, message=FALSE, warning=FALSE}
rm(list=ls())
library("data.table"); library("dplyr"); library("ggplot2")
setwd('~/GitHub/05_ReproducableResearch/RepData_PeerAssessment2')
```
Loading data  
```{r, tidy=TRUE, cache=TRUE}
f <- file.path(getwd(),'2FStormData.csv.bz2')
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'
download.file(url, f, mode="wb") 
DT_storm <- data.table(read.csv(f))
```
Selecting relevant subset
```{r, tidy=TRUE}
dt <- DT_storm %>% select(EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```
Cleaning the data  
```{r, tidy=TRUE}
# Replace empty values with 0's
dt[FATALITIES == "", FATALITIES := "0"]
dt[INJURIES == "", INJURIES := "0"]
dt[PROPDMG == "", PROPDMG := "0"]
dt[CROPDMG == "", CROPDMG := "0"]
# Normalize damage values
dt[PROPDMGEXP == "", PROPDMGEXP := "0"]
dt[PROPDMGEXP == "+" | PROPDMGEXP == "-" | PROPDMGEXP == "?", PROPDMGEXP := "1"]
dt[PROPDMGEXP == "h" | PROPDMGEXP == "H", PROPDMGEXP := "2"]
dt[PROPDMGEXP == "k" | PROPDMGEXP == "K", PROPDMGEXP := "3"]
dt[PROPDMGEXP == "m" | PROPDMGEXP == "M", PROPDMGEXP := "6"]
dt[PROPDMGEXP == "B", PROPDMGEXP := "9"]
dt[CROPDMGEXP == "", CROPDMGEXP := "0"]
dt[CROPDMGEXP == "+" | CROPDMGEXP == "-" | CROPDMGEXP == "?", CROPDMGEXP := "1"]
dt[CROPDMGEXP == "h" | CROPDMGEXP == "H", CROPDMGEXP := "2"]
dt[CROPDMGEXP == "k" | CROPDMGEXP == "K", CROPDMGEXP := "3"]
dt[CROPDMGEXP == "m" | CROPDMGEXP == "M", CROPDMGEXP := "6"]
dt[CROPDMGEXP == "B", CROPDMGEXP := "9"]
# Verify normalized values look ok
dt %>% select(PROPDMGEXP) %>% distinct() %>% setorder(PROPDMGEXP) %>% data.table() %>% invisible()
dt %>% select(CROPDMGEXP) %>% distinct() %>% setorder(CROPDMGEXP) %>% data.table() %>% invisible()
# Recode values as numeric for computation of next step
dt[,PROPDMGEXP := as.integer(PROPDMGEXP)]
dt[,CROPDMGEXP := as.integer(CROPDMGEXP)]
```
## Data manipulations  
Calculating Total Damages for each weather event  
```{r, tidy=TRUE}
dt[, damageUSD := PROPDMG * 10^(PROPDMGEXP) + CROPDMG * 10^(CROPDMGEXP)] 
```  
## Data aggregations  
Calculating total `FATALITIES` by `EVTYPE`  
```{r, tidy=TRUE}
dt <- group_by(dt, EVTYPE) #group data
fatalities <- dt %>% summarize(FATALITIES=sum(FATALITIES))
summarise(ungroup(dt), sum(FATALITIES)) %>% invisible() #ungroup data
```
Calculating total `INJURIES` by `EVTYPE`
```{r, tidy=TRUE}
dt <- group_by(dt, EVTYPE) #group data
injuries <- dt %>% summarize(INJURIES=sum(INJURIES))
summarise(ungroup(dt), sum(INJURIES)) %>% invisible() #ungroup data
```
Calculating total `damageUSD` by `EVTYPE`  
```{r, tidy=TRUE}
dt <- group_by(dt, EVTYPE) #group data
damages <- dt %>% summarize(DAMAGES.USD=sum(damageUSD))
summarise(ungroup(dt), sum(damageUSD)) %>% invisible() #ungroup data
```
## Results  
Identifying top event types by fatalities   
```{r, tidy=TRUE}
topFatalities <- fatalities %>% setorder(-FATALITIES) %>% head(10)
```  
Identifying top event types by injuries  
```{r, tidy=TRUE}
topInjuries <- injuries %>% setorder(-INJURIES) %>% head(10)
```  
Identifying top event types by damages  
```{r, tidy=TRUE}
topDamages <- damages %>% setorder(-DAMAGES.USD) %>% head(10)
```  

### Which types of events are most harmful to population health?  

Plotting total `FATALITIES` by `EVTYPE` for top 10 event types  
```{r, tidy=TRUE}
ggplot(data = topFatalities, 
       aes(x = topFatalities$EVTYPE, y = topFatalities$FATALITIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event type") + 
        ylab("Number of fatalities") + 
        ggtitle("Total Fatalities by Weather Event Type, \n Top 10 Event Types, 1950-2011")
```  

Plotting total `INJURIES` by `EVTYPE` for top 10 event types  
```{r, tidy=TRUE}
ggplot(data = topInjuries, 
       aes(x = topInjuries$EVTYPE, y = topInjuries$INJURIES)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event type") + 
        ylab("Number of injuries") + 
        ggtitle("Total Injuries by Weather Event Type, \n Top 10 Event Types, 1950-2011")
```  

Above we observe that, as measured by fatalities and injuries, Tornados are most harmful to population health.  

Reporting the top event type for fatalities and injuries we get  
```{r}
fatalities %>% setorder(-FATALITIES) %>% head(1) %>% data.table()
injuries %>% setorder(-INJURIES) %>% head(1) %>% data.table()
```  

### Which types of events have the greatest economic consequences?  

Plotting total `DAMAGES.USD` by `EVTYPE` for top 10 event types   
```{r}
ggplot(data = topDamages,
       aes(x = topDamages$EVTYPE, y = topDamages$DAMAGES.USD)) + 
        geom_bar(stat = "identity") + 
        theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
        xlab("Event type") + 
        ylab("Damages (USD)") + 
        ggtitle("Total Damages by Weather Event Type, \n Top 10 Event Types, 1950-2011")
```  

Above we observe that, as measured by damages, Floods have the greatest economic consequences?  

Reporting the top event type for damages we get  
```{r}
damages %>% setorder(-DAMAGES.USD) %>% head(1) %>% data.table()
```  


